---
title: "Metadata_phenotype_QTL2_NF"
author: "Dawson Budke"
date: "2025-08-29"
output: html_document
---

**Purpose**

This documentation serves as the only preparatory step to run the automated "QTL2_NF" Nextflow bioinformatic pipeline. In order to automatically implement a full QTL, from Broman's r/qtl2, pipeline, we first need our project specific input metadata/phenotype data in a combined csv file.

**Overview**

Section 1: Load and merge metadata + phenotypes (physical phenotypes (QTL), transcript counts (eQTL), metabolite counts (mQTL), etc) – input Module 1

Section 2: Download MUGA reference files – input Modules 2/3

Section 3: Download mouse gene database – input Modules 2/3

Section 4: Validate file formats and compatibility

Section 5: Create organized input directory structure

**File Format Requirements**

Row 1:

-   Column A = empty

-   Column B = covariate

-   Column where phenotypes start = phenotype

Row 2:

-   actual column names (SampleID, sex, ngen, batch, transcript/pehno names..)

Remaining rows = sample data

## 

## Load Libraries

```{r}
# Clear environment to start fresh (if necessary)
rm(list = ls())

library(openxlsx)
library(dplyr)
library(tibble)
```

# 1. Load and merge metadata and phenotype data

```{r}
# Load data
Metadata = read.xlsx("DOCholine_AllBatches_Metadata_220202.xlsx")
Phenotypes = read.csv("DOCHLN_combined_counts.csv")

# 382 common samples

###############
# Project-specific modifications
###############

# Step 1: Remove duplicate samples FIRST (using original naming format)
duplicate_ids <- c("323", "368", "371", "372", "374", "376")
duplicate_columns <- paste0("DOCHLN", duplicate_ids)  # Original format without underscore

cat("Original phenotype column names (first 10):\n")
print(colnames(Phenotypes)[1:10])

cat("Looking for duplicate columns to remove:\n")
print(intersect(duplicate_columns, colnames(Phenotypes)))

Phenotypes_clean <- Phenotypes %>% 
 select(-all_of(duplicate_columns))

cat("Removed", length(duplicate_columns), "duplicate samples\n")
cat("Clean phenotypes dimensions:", dim(Phenotypes_clean), "\n")

# Step 2: NOW standardize metadata sample names
cat("Original metadata sample names (first 10):\n")
print(head(Metadata$mouse_ID, 10))

Metadata$mouse_ID <- Metadata$mouse_ID %>%
 gsub("^Dochln", "DOchln", .) %>%
 gsub("^DOCHLN", "DOchln", .) %>%
 gsub("^DOChln", "DOchln", .) %>%
 gsub("^DOchln([0-9])", "DOchln_\\1", .) %>%
 gsub("__", "_", .)

cat("Standardized metadata sample names (first 10):\n")
print(head(Metadata$mouse_ID, 10))

# Step 3: Standardize phenotype column names
cat("Phenotype column names before standardization (first 10):\n")
print(colnames(Phenotypes_clean)[1:10])

colnames(Phenotypes_clean) <- colnames(Phenotypes_clean) %>%
 gsub("^Dochln", "DOchln", .) %>%
 gsub("^DOCHLN", "DOchln", .) %>%
 gsub("^DOChln", "DOchln", .) %>%
 gsub("^DOchln([0-9])", "DOchln_\\1", .) %>%
 gsub("__", "_", .)

cat("Standardized phenotype column names (first 10):\n")
print(colnames(Phenotypes_clean)[1:10])

# Verify standardization worked
metadata_pattern <- unique(substr(Metadata$mouse_ID, 1, 7))
phenotype_pattern <- unique(substr(colnames(Phenotypes_clean)[grep("DOchln", colnames(Phenotypes_clean))], 1, 7))

cat("\nStandardization check:\n")
cat("Metadata prefix patterns:", paste(metadata_pattern, collapse = ", "), "\n")
cat("Phenotype prefix patterns:", paste(phenotype_pattern, collapse = ", "), "\n")

# Step 4: Transpose phenotype data (samples as rows, transcripts as columns)
Phenotypes_transposed <- Phenotypes_clean %>%
 column_to_rownames("X") %>%  
 t() %>%                      
 as.data.frame()

cat("Transposed phenotypes dimensions:", dim(Phenotypes_transposed), "\n")

# Step 5: Check sample ID overlap
metadata_samples <- Metadata$mouse_ID
phenotype_samples <- rownames(Phenotypes_transposed)

common_samples <- intersect(metadata_samples, phenotype_samples)
metadata_only <- setdiff(metadata_samples, phenotype_samples)  
phenotype_only <- setdiff(phenotype_samples, metadata_samples)

cat("Samples in both datasets:", length(common_samples), "\n")
cat("Samples only in metadata:", length(metadata_only), "\n") 
cat("Samples only in phenotypes:", length(phenotype_only), "\n")

# Step 6: Merge datasets keeping only samples present in both
if(length(common_samples) > 0) {
 Metadata_indexed <- Metadata %>% column_to_rownames("mouse_ID")
 Metadata_matched <- Metadata_indexed[common_samples, ]
 Phenotypes_matched <- Phenotypes_transposed[common_samples, ]
 
 stopifnot(identical(rownames(Metadata_matched), rownames(Phenotypes_matched)))
 
 Combined_data <- cbind(Metadata_matched, Phenotypes_matched)
 
 cat("Final combined dataset dimensions:", dim(Combined_data), "\n")
 cat("Number of metadata columns:", ncol(Metadata_matched), "\n")
 cat("Number of transcript columns:", ncol(Phenotypes_matched), "\n")
 cat("First few column names:\n")
 print(colnames(Combined_data)[1:10])
} else {
 cat("ERROR: No overlapping samples found after standardization!\n")
 cat("Sample debugging info:\n")
 cat("First few metadata samples:", head(metadata_samples, 5), "\n")
 cat("First few phenotype samples:", head(phenotype_samples, 5), "\n")
}
```

```{r}
# Step 7: Format for r/qtl2 and Nextflow Module 1
# Define which metadata columns to use as covariates for QTL analysis
covariate_columns <- c("batch", "Sex", "generation", "coat_color", "Diet")  
# Note: Ensure "Sex" and "generation" are included as they're required for r/qtl2 DO crosses

# Check that required covariates exist
required_covars <- c("Sex", "generation")
missing_required <- setdiff(required_covars, covariate_columns)
if(length(missing_required) > 0) {
 cat("WARNING: Missing required covariates for r/qtl2:", paste(missing_required, collapse = ", "), "\n")
}

# Calculate dimensions
n_covariates <- length(covariate_columns)
n_phenotypes <- ncol(Phenotypes_matched)

cat("Formatting data for r/qtl2:\n")
cat("- Samples:", nrow(Combined_data), "\n")
cat("- Covariates:", n_covariates, "\n") 
cat("- Phenotypes (transcripts):", n_phenotypes, "\n")

# Select and reorder columns: covariates first, then all phenotypes
final_data <- Combined_data %>%
 select(all_of(covariate_columns), colnames(Phenotypes_matched)) %>%
 rownames_to_column("SampleID")

# Create properly formatted rows for r/qtl2 CSV structure
# Row 1: Header indicators showing where covariates and phenotypes begin
row1 <- c("", "covariate", rep("", n_covariates-1), "phenotype", rep("", n_phenotypes-1))

# Row 2: Actual column names
row2 <- colnames(final_data)

# Convert data to character matrix to avoid type conflicts
data_matrix <- as.matrix(final_data)
mode(data_matrix) <- "character"

# Combine all rows
all_rows <- rbind(row1, row2, data_matrix)

# Write the formatted CSV file
write.table(all_rows, "formatted_phenotypes_for_nextflow.csv", 
           sep = ",", row.names = FALSE, col.names = FALSE, quote = FALSE)

cat("Created formatted file for Nextflow Module 1\n")
cat("File: formatted_phenotypes_for_nextflow.csv\n")
cat("Dimensions: ", nrow(all_rows), " rows x ", ncol(all_rows), " columns\n")

# Verify file structure
check <- read.csv("formatted_phenotypes_for_nextflow.csv", nrows = 3, header = FALSE, stringsAsFactors = FALSE)
cat("File verification - first 3 rows, first 8 columns:\n")
print(check[1:3, 1:8])

cat("\nReady for Nextflow command:\n")
cat("nextflow run main.nf --phenotype_file formatted_phenotypes_for_nextflow.csv --prefix DOchln\n")
```

QTL2_NF Module 1 input metadata/phenotype CSV is now ready to rock

# 2. Download MUGA reference files
